A full approximation storage multilevel method \\ for nonlinear variational inequalities

### 2023-06-30 paper/fascd.tex
Some edits and removals; we don't appear to need the proof of Lemma 3.3, but I've just commented it out for now

\begin{example}  \label{ex:advectiondiffusion}  Suppose $\partial\Omega$ is partitioned into Dirichlet and Neumann portions, i.e.~$\partial\Omega = \partial_D\Omega \cup \partial_N\Omega$, with $\partial_D\Omega$ of positive measure.  Let $\cV \subset W^{1,2}(\Omega)$ be the space of functions which are zero along $\partial_D\Omega$.  Consider a divergence-free velocity field $\bX$ on $\Omega$ satisfying the conditions of Lemma \ref{lem:advectionskew}, and assume that the flow is outward on the Neumann boundary: $\bX \cdot \bn \ge 0$ on $\partial_N\Omega$.  For $u,v \in \cV$, $\eps>0$, and $\phi \in L^\infty(\Omega)$, define
\begin{equation}
\ip{f(u)}{v} = \eps \left(\grad u, \grad v\right)_{L^2(\Omega)} + b(u,v) - \ip{\phi}{v}. \label{eq:advectiondiffusion}
\end{equation}
\pef{Why is $\phi \in L^\infty(\Omega)$? If that is the case, then should we have an $L^2$ inner product in the last term, rather than a dual pairing?}
With this operator, consider VI \eqref{eq:vi} with $\ell = 0$ for any closed and convex $\cK \subset \cV$. The interior condition, i.e.~PDE strong form, is the following linear advection-diffusion equation
\begin{equation}
-\eps \grad^2 u + \bX\cdot \grad u = \phi.
\label{eq:pdeadvectiondiffusion}
\end{equation}
On the other hand, it is easy to see that $|\ip{f(u)}{v}| \le (\eps + \|\bX\|_\infty + \|\phi\|_\infty) \|u\| \|v\|$, thus that $f:\cK \to \cV'$ is continuous.  Lemma \ref{lem:advectionskew} says that the bilinear form $b(u,v)$ is skew-symmetric up to a nonnegative term.  By the outward flow assumption and the Poincar\'e inequality,
\begin{align*}
\ip{f(u)-f(v)}{u-v} &= \eps \int_\Omega |\grad u - \grad v|^2\dx + b(u-v,u-v) \\
                    &= \eps \int_\Omega |\grad u - \grad v|^2\dx + \frac{1}{2} \int_{\partial_N\Omega} (u-v)^2 \bX\cdot\bn \ge \eps C \|u-v\|^2.
\end{align*}
Thus $f$ is 2-coercive, and so the associated VI problem is well-posed.
\end{example}

@@

In VIs \eqref{eq:cdaddvi} and \eqref{eq:cdmultvi} the argument of $f$ is an element of $\cK$.  Furthermore, by \eqref{eq:constraintdecomp} and \eqref{eq:constraintrestrictionsum} one may write the test vector $v_i - \hat w_i \in \cV_i$ as a difference of admissible vectors from $\cK$.  For \eqref{eq:cdaddvi} and \eqref{eq:cdmultvi} respectively:
\begin{align}
[u - \Pi_i u + v_i] - [u - \Pi_i u + \hat w_i] &= v_i - \hat w_i, \label{eq:admissibledifferenceadd} \\
\left[\sum_{j<i} w_j + v_i + \sum_{j>i} \Pi_j u\right] - \left[\sum_{j<i} w_j + \hat w_i + \sum_{j>i} \Pi_j u\right] &= v_i - \hat w_i.  \label{eq:admissibledifferencemult}
\end{align}
In this sense the VIs over the subsets $\mathcal{K}_i$ solve the same problem as the original one.

@@

We call \pr{cd-add} and \pr{cd-mult} the \emph{basic CD iterations}.  VI problems \eqref{eq:cdaddvi} and \eqref{eq:cdmultvi} reduce to constrained line searches when the $\cV_i$ are one-dimensional.  As shown in the next Lemma, when VI problem \eqref{eq:vi} corresponds to convex optimization \eqref{eq:minimization} then, subject to bounds on the damping $\alpha$, these iterations generate monotonically-decreasing objective values.

\begin{lemma} \cite{Tai2003}  Suppose $f=F'$ for $F:\cK\to\RR$ convex and differentiable, and let $u\in\cK$.  For $0 \le \alpha \le 1$ the output $w \in \cK$ from \pr{cd-mult} satisfies
\begin{equation}
F(w) - \ip{\ell}{w} \le F(u) - \ip{\ell}{u}.  \label{eq:objectivemonotone}
\end{equation}
If $0 \le \alpha \le 1/m$ then \pr{cd-add} likewise satisfies \eqref{eq:objectivemonotone}.
\end{lemma}

\begin{proof}  We assume $\ell=0$ without loss of generality; otherwise define $\tilde F(u)=F(u) - \ip{\ell}{u}$ and observe that the proof goes through for $\tilde F$ essentially unchanged.

To show that each iteration of the {for} loop in \pr{cd-mult} lowers the objective value, let $z_{i} = \sum_{j\le i} w_j + \sum_{j > i} \Pi_j u \in \cK$, $i=0,\dots,m-1$; this is the state at the end of the $i$th iteration.  Note that $z_{i-1} = \sum_{j < i} w_j + \Pi_i u + \sum_{j > i} \Pi_j u$ for $i>0$.  Therefore by convexity of $F$, observation \eqref{eq:admissibledifferencemult} with $v_i=\Pi_i u$, and \eqref{eq:cdmultvi} we find:
\begin{equation*}
F(z_{i-1}) - F\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right) \ge \ip{f\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right)}{\Pi_i u - \hat w_i} \ge 0.
\end{equation*}
Using convexity (again) now gives
\begin{align*}
F(z_i) &= F\left((1-\alpha) z_{i-1} + \alpha \left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right)\right) \\
&\le  (1-\alpha) F(z_{i-1}) + \alpha F\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right) \le F(z_{i-1}).
\end{align*}
As $u=z_0$ and $w = z_{m-1}$, the result for \pr{cd-mult} follows.

Suppose $w=u+\alpha\sum_i e_i$ is computed by \pr{cd-add} using $\alpha \le 1/m$.  Considering each $e_i=\hat w_i - \Pi_i u$ separately, convexity of $F$ and VI \eqref{eq:cdaddvi} with $v_i=\Pi_i u$ yields
\begin{equation}
F(u) - F(u+e_i) \ge \ip{f(u+e_i)}{u-(u+e_i)} = \ip{f(u-\Pi_i u + \hat w_i)}{\Pi_i u - \hat w_i} \ge 0. \label{eq:cdaddonereduction}
\end{equation}
Let $v = u+ \frac{1}{m} \sum_i e_i$.  Again using convexity (Jensen's inequality), and \eqref{eq:cdaddonereduction}, we have
\begin{equation*}
F(v) = F\left(\frac{1}{m} \sum_{i=0}^{m-1} (u+e_i)\right) \le \frac{1}{m} \sum_{i=0}^{m-1} F(u+e_i) \le \frac{1}{m} \sum_{i=0}^{m-1} F(u) = F(u).
\end{equation*}
Now define $\lambda = m\alpha$ so $0 \le \lambda \le 1$ and $w = (1-\lambda) u + \lambda v$.  Thus by convexity and the above result, $F(w) \le (1-\lambda) F(u) + \lambda F(v) \le F(u)$.
\end{proof}

When $\alpha=1/m$ then the \pr{cd-add} proof can be simplified because $v=w$ and $\lambda=1$.  In this case $F(w)$ is less than $\frac{1}{m} \sum_i F(u+e_i)$ by Jensen's inequality, which is lower than $F(u)$ by \eqref{eq:cdaddvi} or \eqref{eq:cdmultvi}, which gives inequality \eqref{eq:cdaddonereduction}.

The basic CD iterations are meaningful whether or not we are in the optimization case $f=F'$, and even when $f$ is nonlinear, non-local, or defined only on $\cK$.  However, practical and efficient FE implementation for such general operators $f$ seems not to have been addressed in the literature.  References \cite{GraeserKornhuber2009,Tai2003} only apply the CD method to the classical obstacle problem, wherein $f=F'$ is linear, local, and defined for all inputs (on all of $\mathcal{V}$).  Substantial implementation simplifications are available in such cases.

\begin{example}  \label{ex:fnice} Suppose $f:\cV \to \cV'$ is linear and a differential operator, thus local in the sense that for any $z\in\mathcal{V}$ the value $\ip{f(\phi)}{z}$ can be computed by an integral over the support of $\phi$.  Considering \pr{cd-add} for concreteness, by linearity \eqref{eq:cdaddvi} can be rewritten
\begin{equation}
\ip{f(e_i)}{v_i-\hat w_i} \ge \ip{\tilde\ell}{v_i-\hat w_i}, \label{eq:linearlocalvi}
\end{equation}
for all $v_i \in \mathcal{K}_i$, where $\tilde\ell = \ell - f(u)$ incorporates the initial residual $f(u)$.  Note $e_i = \hat w_i - \Pi_i u \in \cV_i$ is not generally admissible---generally $e_i \notin \cK$---but \eqref{eq:linearlocalvi} is meaningful for this $f$.
\end{example}

@@

From now on we will restrict to the case of \emph{box constraints} \cite{BensonMunson2006,FerrisPang1997}, equivalently \emph{bi-obstacle problems}.  Suppose $\Omega \subset \RR^d$ is open and let $\mathcal{V}=W^{1,p}(\Omega)$.  The \emph{lower obstacle} $\underline{\gamma}$ and the \emph{upper obstacle} $\overline{\gamma}$ are given functions, defined on the closure $\overline{\Omega}$, with values in the extended real line $\tilde\RR = [-\infty,+\infty]$.  We suppose $\underline{\gamma} : \overline{\Omega} \to [-\infty,+\infty)$ and $\overline{\gamma} : \overline{\Omega} \to (-\infty,+\infty]$ are measurable, $\underline{\gamma} \le \overline{\gamma}$ on $\overline{\Omega}$, and $\underline{\gamma} \le g_D \le \overline{\gamma}$ on $\partial_D \Omega$, and that $\partial\Omega$ is split into disjoint, measurable sets, $\partial\Omega = \partial_D \Omega \cup \partial_N \Omega$, with $\partial_D \Omega$ of positive measure, to facilitate well-posedness \cite{Evans2010}, and $\partial_N \Omega$ possibly empty.  Dirichlet data\footnote{Dirichlet boundary values are in the trace sense \cite{Evans2010}, and $g_D$ is assumed to be as regular as needed.} $g_D:\partial_D \Omega \to \RR$ then defines
\begin{equation}
\cK = \left\{w\,:\,\underline{\gamma} \le w \le \overline{\gamma} \, \text{ and }\, w\big|_{\partial_D \Omega} = g_D\right\} \subset \cV =W^{1,p}(\Omega), \label{eq:originalconstraintset}
\end{equation}
the closed and convex \emph{constraint set}.  Note that we are treating inequality constraints and boundary conditions in a unified manner.\footnote{Any Neumann conditions over $\partial_N \Omega$ are determined by the operator formula for $f$, i.e.~by an integration over $\partial_N\Omega$.}

Recalling abstract VI problem \eqref{eq:vi}, from now on we only constraint sets \eqref{eq:originalconstraintset}.  Given continuous $f:\cK \to \cV'$, and $\ell \in \cV'$, we will assume that the following is well-posed for $u^\star\in \cK$:
\begin{equation}
\ip{f(u^\star)}{v-u^\star} \ge \ip{\ell}{v-u^\star} \qquad \text{for all } v\in \cK. \label{eq:boxdirichletvi}
\end{equation}
Note that $v-u^\star \in W_0^{1,p}(\Omega)$.

>>>>

\begin{example}  \label{ex:advectiondiffusion}  Suppose $\partial\Omega$ is partitioned into Dirichlet and Neumann portions, i.e.~$\partial\Omega = \partial_D\Omega \cup \partial_N\Omega$, with $\partial_D\Omega$ of positive Hausdorff measure.  Let $\cV \subset W^{1,2}(\Omega)$ be the space of functions which are zero along $\partial_D\Omega$.  Consider a divergence-free velocity field $\bX$ on $\Omega$ satisfying the conditions of Lemma \ref{lem:advectionskew}, and assume that the flow is outward on the Neumann boundary: $\bX \cdot \bn \ge 0$ on $\partial_N\Omega$.  For $u,v \in \cV$, $\eps>0$, and $\phi \in L^\infty(\Omega)$, define
\begin{equation}
\ip{f(u)}{v} = \eps \left(\grad u, \grad v\right)_{L^2(\Omega)} + b(u,v) - \ip{\phi}{v}. \label{eq:advectiondiffusion}
\end{equation}
\pef{Why is $\phi \in L^\infty(\Omega)$? If that is the case, then should we have an $L^2$ inner product in the last term, rather than a dual pairing?}
With this operator, consider VI \eqref{eq:vi} with $\ell = 0$ for any closed and convex $\cK \subset \cV$. The interior condition, i.e.~PDE strong form, is the following linear advection-diffusion equation
\begin{equation}
-\eps \grad^2 u + \bX\cdot \grad u = \phi.
\label{eq:pdeadvectiondiffusion}
\end{equation}
On the other hand, it is easy to see that $|\ip{f(u)}{v}| \le (\eps + \|\bX\|_\infty + \|\phi\|_\infty) \|u\| \|v\|$, thus that $f:\cK \to \cV'$ is continuous.  Lemma \ref{lem:advectionskew} says that the bilinear form $b(u,v)$ is skew-symmetric up to a nonnegative term.  By the outward flow assumption and the Poincar\'e inequality,
\begin{align*}
\ip{f(u)-f(v)}{u-v} &= \eps \int_\Omega |\grad u - \grad v|^2\dx + b(u-v,u-v) \\
                    &= \eps \int_\Omega |\grad u - \grad v|^2\dx + \frac{1}{2} \int_{\partial_N\Omega} (u-v)^2 \bX\cdot\bn \ge \eps C \|u-v\|^2.
\end{align*}
Thus $f$ is 2-coercive, and so the associated VI problem is well-posed.
\end{example}

@@

%In VIs \eqref{eq:cdaddvi} and \eqref{eq:cdmultvi} the argument of $f$ is an element of $\cK$.  Furthermore, by \eqref{eq:constraintdecomp} and \eqref{eq:constraintrestrictionsum} one may write the test vector $v_i - \hat w_i \in \cV_i$ as a difference of admissible vectors from $\cK$.  For \eqref{eq:cdaddvi} and \eqref{eq:cdmultvi} respectively:
%\begin{align}
%[u - \Pi_i u + v_i] - [u - \Pi_i u + \hat w_i] &= v_i - \hat w_i, \label{eq:admissibledifferenceadd} \\
%\left[\sum_{j<i} w_j + v_i + \sum_{j>i} \Pi_j u\right] - \left[\sum_{j<i} w_j + \hat w_i + \sum_{j>i} \Pi_j u\right] &= v_i - \hat w_i.  \label{eq:admissibledifferencemult}
%\end{align}
%In this sense the VIs over the subsets $\mathcal{K}_i$ solve the same problem as the original one.

@@

We call \pr{cd-add} and \pr{cd-mult} the \emph{basic CD iterations}.  VI problems \eqref{eq:cdaddvi} and \eqref{eq:cdmultvi} reduce to constrained line searches when the $\cV_i$ are one-dimensional.  As shown in the next Lemma, when VI problem \eqref{eq:vi} corresponds to convex optimization \eqref{eq:minimization} then, subject to bounds on the damping $\alpha$, these iterations generate monotonically-decreasing objective values.
\begin{lemma} \cite{Tai2003}  Suppose $f=F'$ for $F:\cK\to\RR$ convex and differentiable, and let $u\in\cK$.  For $0 \le \alpha \le 1$ the output $w \in \cK$ from \pr{cd-mult} satisfies
\begin{equation}
F(w) - \ip{\ell}{w} \le F(u) - \ip{\ell}{u}.  \label{eq:objectivemonotone}
\end{equation}
If $0 \le \alpha \le 1/m$ then \pr{cd-add} likewise satisfies \eqref{eq:objectivemonotone}.
\end{lemma}

%\begin{proof}  We assume $\ell=0$ without loss of generality; otherwise define $\tilde F(u)=F(u) - \ip{\ell}{u}$ and observe that the proof goes through for $\tilde F$ essentially unchanged.
%
%To show that each iteration of the {for} loop in \pr{cd-mult} lowers the objective value, let $z_{i} = \sum_{j\le i} w_j + \sum_{j > i} \Pi_j u \in \cK$, $i=0,\dots,m-1$; this is the state at the end of the $i$th iteration.  Note that $z_{i-1} = \sum_{j < i} w_j + \Pi_i u + \sum_{j > i} \Pi_j u$ for $i>0$.  Therefore by convexity of $F$, observation \eqref{eq:admissibledifferencemult} with $v_i=\Pi_i u$, and \eqref{eq:cdmultvi} we find:
%\begin{equation*}
%F(z_{i-1}) - F\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right) \ge \ip{f\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right)}{\Pi_i u - \hat w_i} \ge 0.
%\end{equation*}
%Using convexity (again) now gives
%\begin{align*}
%F(z_i) &= F\left((1-\alpha) z_{i-1} + \alpha \left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right)\right) \\
%&\le  (1-\alpha) F(z_{i-1}) + \alpha F\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right) \le F(z_{i-1}).
%\end{align*}
%As $u=z_0$ and $w = z_{m-1}$, the result for \pr{cd-mult} follows.
%
%Suppose $w=u+\alpha\sum_i e_i$ is computed by \pr{cd-add} using $\alpha \le 1/m$.  Considering each $e_i=\hat w_i - \Pi_i u$ separately, convexity of $F$ and VI \eqref{eq:cdaddvi} with $v_i=\Pi_i u$ yields
%\begin{equation}
%F(u) - F(u+e_i) \ge \ip{f(u+e_i)}{u-(u+e_i)} = \ip{f(u-\Pi_i u + \hat w_i)}{\Pi_i u - \hat w_i} \ge 0. \label{eq:cdaddonereduction}
%\end{equation}
%Let $v = u+ \frac{1}{m} \sum_i e_i$.  Again using convexity (Jensen's inequality), and \eqref{eq:cdaddonereduction}, we have
%\begin{equation*}
%F(v) = F\left(\frac{1}{m} \sum_{i=0}^{m-1} (u+e_i)\right) \le \frac{1}{m} \sum_{i=0}^{m-1} F(u+e_i) \le \frac{1}{m} \sum_{i=0}^{m-1} F(u) = F(u).
%\end{equation*}
%Now define $\lambda = m\alpha$ so $0 \le \lambda \le 1$ and $w = (1-\lambda) u + \lambda v$.  Thus by convexity and the above result, $F(w) \le (1-\lambda) F(u) + \lambda F(v) \le F(u)$.
%\end{proof}
%When $\alpha=1/m$ then the \pr{cd-add} proof can be simplified because $v=w$ and $\lambda=1$.  In this case $F(w)$ is less than $\frac{1}{m} \sum_i F(u+e_i)$ by Jensen's inequality, which is lower than $F(u)$ by \eqref{eq:cdaddvi} or \eqref{eq:cdmultvi}, which gives inequality \eqref{eq:cdaddonereduction}.

The basic CD iterations are meaningful whether or not we are in the optimization case $f=F'$, and even when $f$ is nonlinear, non-local, or defined only on $\cK$.  However, practical and efficient FE implementation for such general operators $f$ seems not to have been addressed in the literature.  References \cite{GraeserKornhuber2009,Tai2003} only apply the CD method to the classical obstacle problem, wherein $f=F'$ is linear, local, and defined on all of $\mathcal{V}$. Substantial implementation simplifications are available in such cases.

\begin{example}  \label{ex:fnice} Suppose $f:\cV \to \cV'$ is linear and is local in the sense that for any $z\in\mathcal{V}$ the value $\ip{f(\phi)}{z}$ can be computed by an integral over the support of $\phi$.  Considering \pr{cd-add} for concreteness, by linearity \eqref{eq:cdaddvi} can be rewritten
\begin{equation}
\ip{f(e_i)}{v_i-\hat w_i} \ge \ip{\tilde\ell}{v_i-\hat w_i}, \label{eq:linearlocalvi}
\end{equation}
for all $v_i \in \mathcal{K}_i$, where $\tilde\ell = \ell - f(u)$ incorporates the initial residual $f(u)$.  Note $e_i = \hat w_i - \Pi_i u \in \cV_i$ is not generally admissible---generally $e_i \notin \cK$---but \eqref{eq:linearlocalvi} is meaningful for this $f$.
\pef{Can you expand a little bit here? I confess I don't follow.}
\end{example}

@@

From now on we restrict to the case of \emph{box constraints} \cite{BensonMunson2006,FerrisPang1997}, equivalently \emph{bi-obstacle problems}.  Suppose $\Omega \subset \RR^d$ is open and let $\mathcal{V}=W^{1,p}(\Omega)$.  The \emph{lower obstacle} $\underline{\gamma}$ and the \emph{upper obstacle} $\overline{\gamma}$ are given functions, defined on the closure $\overline{\Omega}$, with values in the extended real line $\tilde\RR = [-\infty,+\infty]$.  We suppose $\underline{\gamma} : \overline{\Omega} \to [-\infty,+\infty)$ and $\overline{\gamma} : \overline{\Omega} \to (-\infty,+\infty]$ are measurable, $\underline{\gamma} \le \overline{\gamma}$ on $\overline{\Omega}$, and $\underline{\gamma} \le g_D \le \overline{\gamma}$ on $\partial_D \Omega$, and that $\partial\Omega$ is split into disjoint, measurable sets, $\partial\Omega = \partial_D \Omega \cup \partial_N \Omega$, with $\partial_D \Omega$ of positive Hausdorff measure, and $\partial_N \Omega$ possibly empty.  Dirichlet data\footnote{Dirichlet boundary values are in the trace sense \cite{Evans2010}, and $g_D$ is assumed to be as regular as needed.} $g_D:\partial_D \Omega \to \RR$ then defines
\begin{equation}
\cK = \left\{w\,:\,\underline{\gamma} \le w \le \overline{\gamma} \, \text{ and }\, w\big|_{\partial_D \Omega} = g_D\right\} \subset \cV =W^{1,p}(\Omega), \label{eq:originalconstraintset}
\end{equation}
the closed and convex \emph{constraint set}.  Note that we are treating inequality constraints and boundary conditions in a unified manner.\footnote{Any Neumann conditions over $\partial_N \Omega$ are imposed weakly in the definition of $f$, i.e.~by an integration over $\partial_N\Omega$.}

We recall the abstract VI problem \eqref{eq:vi} with constraint sets \eqref{eq:originalconstraintset}.  Given continuous $f:\cK \to \cV'$, and $\ell \in \cV'$, we will assume that the following is well-posed for $u^\star\in \cK$:
\begin{equation}
\ip{f(u^\star)}{v-u^\star} \ge \ip{\ell}{v-u^\star} \qquad \text{for all } v\in \cK. \label{eq:boxdirichletvi}
\end{equation}
Note that $v-u^\star \in W_0^{1,p}(\Omega)$.
